---
title: "gapminder-wrangle.rmd"
author: "jenny"
date: "1/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This exercise is to practice wrangling gapminder data usng dplyr.

```{r}
library(tidyverse)   ## install.packages("tidyverse")
```

```{r}
## read gapminder csv. Note the readr:: prefix identifies which package it's in
gapminder <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')
```


```{r}
## explore the gapminder dataset
gapminder # this is super long! Let's inspect in different ways
```

```{r}
head(gapminder) # shows first 6
tail(gapminder) # shows last 6

head(gapminder, 10) # shows first X that you indicate
tail(gapminder, 12) # guess what this does!
```

```{r}
str(gapminder) # ?str - displays the structure of an object
```

```{r}
names(gapminder)
dim(gapminder)    # ?dim dimension
ncol(gapminder)   # ?ncol number of columns
nrow(gapminder)   # ?nrow number of rows
```

```{r}
install.packages('skimr')
```


```{r}
summary(gapminder)

library(skimr) # install.packages('skimr')
skim(gapminder)
```

```{r}
gapminder$lifeExp # very long! hard to make sense of...
head(gapminder$lifeExp) # can do the same tests we tried before
str(gapminder$lifeExp) # it is a single numeric vector
summary(gapminder$lifeExp) # same information, formatted slightly differently
```

## dplyr

filter() subset observations by values (rows)
select() subset variables by names (columns)
mututate() create new variables with functions of existing (columns)
summarise() collpase values to a summary (rows and columns)
arrange () reorder/sort rows

group_by() to apply functions group-by-group throughout entire dataset

```{r}
filter(gapminder, lifeExp < 29)
```

```{r}
filter(gapminder, country == "Mexico")
```

```{r}
filter(gapminder, country %in% c("Mexico", "Peru"))
```

```{r}
filter(gapminder, country == "Mexico", year == 2002)
```

```{r}
x <- filter(gapminder, country == "Brazil", year > 1987, year <= 2007)
mean(x$lifeExp)
```

```{r}
select(gapminder, year, country, lifeExp)
select(gapminder, -continent, -lifeExp) # you can use - to deselect columns
```

```{r}
## to combine operations and to use syntax that better matches the logic, use the "Pipe" operator %>% 
## For example, the new variable gap_cambodia "gets" the following: start with the gapminder dataset, then filter records for Cambodia, then select all the variables except continent and lifeExp
gap_cambodia  <- gapminder %>% 
  filter(country == "Cambodia") %>%
  select(-continent, -lifeExp) 
```

```{r}
gapminder %>%
  mutate(gdp = pop * gdpPercap)
```

```{r}
gapminder %>% 
  filter (continent == "Asia", year ==2007) %>% 
  mutate (asian_pop = pop/1000)
```

```{r}
gapminder %>%
  filter(year == 2002) %>%
  group_by(continent) %>% 
  mutate(cont_pop = sum(pop))
```

```{r}
gapminder %>%
  group_by(continent) %>%
  summarize(cont_pop = sum(pop)) %>%
  ungroup()
## always remember to ungroup after grouping a variable
```

```{r}
gapminder %>%
  group_by(continent, year) %>%
  summarize(cont_pop = sum(pop))
```

```{r}
gapminder %>%
  group_by(continent, year) %>%
  summarize(cont_pop = sum(pop)) %>%
  arrange(year)
```

```{r}
## my answer: "What is the max GDP per continent across all years?"
gapminder %>%
  mutate(gdp = pop * gdpPercap) %>% 
  group_by(continent) %>% 
  summarize (max_gdp = max(gdp))
```

```{r}
## Their answer
gapminder %>%
  mutate(gdp = pop * gdpPercap) %>%
  group_by(continent) %>%
  mutate(max_gdp = max(gdp)) %>%
  filter(gdp == max_gdp)
```

```{r}
## my answer "What is the maximum life expectancy for each of the countries in Asia - order the countries from lowest to highest."
gapminder %>% 
  filter (continent == "Asia") %>% 
  group_by(country) %>% 
  summarize(lifeExp = max(lifeExp)) %>% 
  arrange(lifeExp)
```

```{r}
## their answer
asia_life_exp <- gapminder %>%
  filter(continent == 'Asia') %>%
  group_by(country) %>%
  filter(lifeExp == max(lifeExp)) %>%
  arrange(year)
asia_life_exp
```

